Reasonably Programmable Literal Notation

### 2018-06-18 icfp18/syntax-icfp18.tex
edits to 2.1.1 and tweaks to spacing

% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}. For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
\begin{figure}[t]
\begin{subfigure}[t]{0.45\textwidth}
\vspace{-8px}
\begin{lstlisting}[mathescape=~]
  module Regex = {
    type t = Empty
           | AnyChar 
           | Str(string)
           | Seq(t, t) 
           | Or(t, t) 
           | Star(t);
  }
\end{lstlisting}
\vspace{-5px}
\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
\label{fig:Regex-module-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.53\textwidth}
\vspace{-8px}
\begin{lstlisting}[mathescape=|]
  module RegexNotation = {
    notation $regex at Regex.t {
      lexer  RegexLexer
      parser RegexParser.start
      in package regex_parser;
      dependencies = {module Regex = Regex}
    }
  }
\end{lstlisting}
\vspace{-5px}
\caption{The definition of the \li{$regex} TLM. Figure \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}.}
\label{fig:regex-tlm-def}
\end{subfigure}

\begin{subfigure}[t]{\textwidth}
\vspace{4px}
\begin{lstlisting}
  notation $regex = RegexNotation.$regex; /* or open RegexNotation */
  module DNA = { let any_base = $regex `(SURLA|T|G|CEURL)`; };
  let bisA = $regex `(SURLGC$(EURLDNA.any_baseSURL)GCEURL)`;
  let restriction_template = (gene) => 
    $regex `(SURL$(EURLbisASURL)$(EURLDNA.any_baseSURL)*$$(EURLgeneSURL)$(EURLDNA.any_baseSURL)*$(EURLbisASURL)EURL)`;
\end{lstlisting}
\vspace{-6px}
\caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
}
\label{fig:first-tlm-example}
\end{subfigure}
\vspace{1px}
\caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
% \vspace{-4px}
\label{fig:regex-case-study}
\end{figure}

@@

\subsubsection{Responsibility} 
Responsibility for lexing, parsing and expanding each literal body is delegated uniquely to the applied TLM. % Choices about how spliced expressions are recognized inside the literal body are entirely at the discretion of each TLM. In this case, the TLM provider has chosen to use \li{$(}$e$\li{)} for regex splicing and \li{$$(}$e$\li{)} for string splicing, where $e$ is an unexpanded Reason expression of arbitrary form. % We return to these topics below.% reasoning about splicing works when encountering an unfamiliar literal form below.
% and take the perspective of a client that does not want to delve into these details.
TLM definitions follow the same scoping rules as Reason modules, i.e. they can be defined in modules alongside other definitions and accessed through module paths. When a TLM definition appears inside a module with an explicitly specified module type, it must also appear in the module type with the same specification, up to the usual notions of type and module path equivalence in the type annotation and module dependencies, which are discussed below. This is much like the situation with datatype definitions in ML. In Sec. \ref{sec:implementation}, we will describe how we use an encoding of TLM definitions as modules with singleton signatures to avoid having to primitively extend OCaml's module system.

What is fundamental about this design is that there is a well-defined protocol for finding the definition of the TLM uniquely responsible for each generalized literal form in a program, following the usual scoping rules of the language. An editor service or documentation tool can use this protocol to integrate TLM definition lookup into a ``go to definition'' command. Moreover, clients can define scoped abbreviations as shown on Line 1 of Fig.~\ref{fig:first-tlm-example}, again following the usual scoping rules. By convention, we define TLMs in a module suffixed with \li{Notation} so that client programmers can \li{open} just the relevant TLM definition(s) without bringing other definitions into scope. We will demonstrate a simple lexically scoped implicit application mechanism for situations where the same TLM is being repeatedly applied  in Sec. \ref{sec:regexparser}.

@@

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
   This work was
supported in part by AFRL and DARPA under agreement \#FA8750-16-2-0042;
 by the NSA under lablet contract \#H98230-14-C-0140; and by Ravi Chugh
 at the University of Chicago. The authors
would like to thank Robert Harper, Karl Crary and Eric Van Wyk for 
valuable feedback on the first author's doctoral dissertation, which forms the basis of this work; 
Charles Chamberlain, 
for his crucial contributions to the implementation and for valuable feedback on the
paper; and members of the PL Reading Group at University of Chicago, and the anonymous referees
at ICFP and prior conferences, for their valuable critiques, which substantially improved the paper.

>>>>

% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}. For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
\begin{figure}[t]
\begin{subfigure}[t]{0.45\textwidth}
\vspace{-4px}
\begin{lstlisting}[mathescape=~]
  module Regex = {
    type t = Empty
           | AnyChar 
           | Str(string)
           | Seq(t, t) 
           | Or(t, t) 
           | Star(t);
  }
\end{lstlisting}
\vspace{-5px}
\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
\label{fig:Regex-module-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.53\textwidth}
\vspace{-4px}
\begin{lstlisting}[mathescape=|]
  module RegexNotation = {
    notation $regex at Regex.t {
      lexer  RegexLexer
      parser RegexParser.start
      in package regex_parser;
      dependencies = {module Regex = Regex}
    }
  }
\end{lstlisting}
\vspace{-5px}
\caption{The definition of \li{$regex}. Fig. \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}, which are detailed in Sec.~\ref{sec:provider-perspective}.}
\label{fig:regex-tlm-def}
\end{subfigure}

\begin{subfigure}[t]{\textwidth}
\vspace{7px}
\begin{lstlisting}
  notation $regex = RegexNotation.$regex; /* or open RegexNotation */
  module DNA = { let any_base = $regex `(SURLA|T|G|CEURL)`; };
  let bisA = $regex `(SURLGC$(EURLDNA.any_baseSURL)GCEURL)`;
  let restriction_template = (gene) => 
    $regex `(SURL$(EURLbisASURL)$(EURLDNA.any_baseSURL)*$$(EURLgeneSURL)$(EURLDNA.any_baseSURL)*$(EURLbisASURL)EURL)`;
\end{lstlisting}
\vspace{-6px}
\caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
}
\label{fig:first-tlm-example}
\end{subfigure}
\vspace{1px}
\caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
% \vspace{-4px}
\label{fig:regex-case-study}
\end{figure}

@@

\subsubsection{Responsibility} 
Responsibility for lexing, parsing and expanding each literal body is delegated uniquely to the applied TLM. % Choices about how spliced expressions are recognized inside the literal body are entirely at the discretion of each TLM. In this case, the TLM provider has chosen to use \li{$(}$e$\li{)} for regex splicing and \li{$$(}$e$\li{)} for string splicing, where $e$ is an unexpanded Reason expression of arbitrary form. % We return to these topics below.% reasoning about splicing works when encountering an unfamiliar literal form below.
% and take the perspective of a client that does not want to delve into these details.
TLM definitions and abbreviations (like the abbreviation on Line 1 of Fig.~\ref{fig:first-tlm-example}) follow the same scoping rules as Reason modules, i.e. they can appear in modules and be accessed through module paths. When a TLM definition appears inside a module with an explicitly specified module type (a.k.a. signature), it must also appear in the module type with the same specification, up to the usual notions of type and module path equivalence in the type annotation and dependencies, which are discussed below. This is much like the situation with datatype definitions in ML. By convention, we define TLMs in a module suffixed with \li{Notation} so that client programmers can \li{open} just the relevant TLM definition(s) without bringing other definitions into scope. We will demonstrate a simple lexically scoped implicit application mechanism for situations where the same TLM is being repeatedly applied  in Sec. \ref{sec:regexparser}.

What is fundamental about this design is that there is a well-defined protocol that follows the usual scoping rules of the language for finding the definition of the TLM uniquely responsible for each generalized literal form in a program. An editor service or documentation tool could use this protocol to integrate TLM definition lookup into a ``go to definition'' command. In Sec. \ref{sec:implementation}, we will describe how we use an encoding of TLM definitions as modules with singleton signatures to avoid having to primitively extend OCaml.

@@

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
   This work was
supported in part by AFRL and DARPA under agreement \#FA8750-16-2-0042;
 by the NSA under lablet contract \#H98230-14-C-0140; and by Ravi Chugh
 at the University of Chicago. The authors
would like to thank Robert Harper, Karl Crary and Eric Van Wyk for 
valuable feedback on the first author's doctoral dissertation, which forms the basis of this work; 
Charles Chamberlain 
for his crucial contributions to the implementation and for valuable feedback on the
paper; and members of the PL Reading Group at University of Chicago and the anonymous referees
at ICFP and prior conferences for their valuable critiques, which substantially improved the paper.
